Discrete Quantum Theories and Computing

### 2018-12-16 dissertation.tex
dissertation.lyx "Chapter 1 Introduction" Updating...

\usepackage{tabularx}

@@

Since we as human cannot distinguish quantities differing only by
an arbitrarily small extent \cite{Turing_1937,Gisin2017}, the quantum
mechanical model based on the continuum of complex numbers may not
capture our ability of predicting quantum phenomenon and the power
of realistic quantum computers. To incorporate our limitation of distinguishability,
but agnostic about whether the reality is ultimately discrete or continuous,
we considered two types of quantum models. By assuming the quantum
states are ultimately discrete and distinguishable, our first type
of quantum theories replaces the complex numbers by finite fields.
We then focus on the limitation of distinguishability itself, incorporate
the finite precision measurement into our model, and consider quantum
interval-valued probability measures.

After first suggested by Richard Feynman 40 years ago~\cite{Feynman1982Simulating},
IBM, \textitalian{Rigetti}, Google, Microsoft, and many other companies
and countries start to build realistic quantum computers recently
\cite{Shieber2018,Kahn2019}, and seek more efficient ways to simulate
chemical molecules and to break down RSA \cite{boneh1999twenty,wiki:RSAFactoring}.
No matter whether we use classical or quantum computers to solve these
two tasks, they require two different types of quantities. The question
and answer of RSA are both discrete integers, but it is beyond our
ability to distinguish whether the quantities required by simulating
molecules are discrete or not. Before we dive into the discussion
of these quantities in quantum computation, it is instructive to review
how classical quantities are characterized and computed.

@@

In contrast, the non-apparently discrete quantities confused many
cultures since very beginning. For example, the ancient Chinese philosopher
Chuang-tzu (庄子) didn't believe these quantities have a minimum unit,
and expressed ``If a one-foot-long stick is cut into halves every
day, the cutting will never come to an end''\footnote{``一尺之棰，日取其半，萬世不竭。'' in Chinese \cite{Zhuangzi1999,Liu2018}.}.
As another example, Greek mathematicians in the school of Pythagoras
believed every mathematical model must be ultimately discrete, and
they in legend murdered Hippasus because he proved there is no minimum
unit between a side of a square and its diagonal. Comparing the idea
that the length of a square's diagonal is not a physical quantity
or there is no square physically, working with a mathematical model
without minimum unit seems more appealing. Since the mathematical
model is continuous, their corresponding physical quantities are naturally
assumed to be continuous. Later, these continuous quantities becomes
the basic component of analog computers. After logarithm was discovered,
a slide rule, also known as a slipstick, is used to compute multiplication,
division, and more complex operations as a mechanical analog computer~\cite{wiki:SlideRule}.
Charge amplifiers are used to build electrical analog computers and
compute the integration in calculus \cite{wiki:Integrator,wiki:ChargeAmplifier}.

Although assuming physical quantities is as continuous as their mathematical
models are convenience, the physical quantities discussed in previous
paragraph are not simply real numbers. Chuang-tzu's stick are made
by molecules and cannot be cut in halves endlessly. The input charge
of an integrator an must actually be an integer multiple of the elementary
charge. A prefect square cannot be draw physically. It is beyond our
ability to decide whether a prefect square exists in the space because
whether space itself is ultimately discrete or continuous is still
an solved question. Even if it exists, human have no ability to distinguish
them from an almost prefect square differing only by an arbitrarily
small extent. Similar situation applying to the length of a slide
rule, it is beyond our ability to decide whether its length is ultimately
discrete or continuous. Even if it is continuous, there is no way
to identify which real number represents the length of a slide rule.
As the results, even if analog computers really store real numbers,
they cannot be read, write, and use to its exact value to branch the
computation. The opposite of the last condition sometimes is assumed
by real computational models. For example, Blum, Cucker, Shub, and
Smale modeled their BCSS machine to allow the operations of deciding
whether a number is greater or equal to zero over real numbers and
whether a number is exactly zero over complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}.
Therefore, the computational power of analog computers is far weaker
than their theoretical prediction.

\begin{table}
\caption{\label{tab:Comparison}Comparison between the classical quantities
and their mathematical models}

\centering{}%
\begin{tabular}{cccc}
\toprule 
\addlinespace
Physical quantities & Discrete & %
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Discrete with\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
extreme small unit\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & We don't know\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational devices & Digital computer & Analog computer & Analog computer\tabularnewline\addlinespace
\midrule
\addlinespace
\addlinespace
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Mathematical\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
representation\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & Integers & Real numbers & Real numbers\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational model  & Turing machine & BCSS machine & BCSS machine\tabularnewline\addlinespace
\midrule
\addlinespace
\addlinespace
How the models realize? & Reliably & Not reliably & Not reliably\tabularnewline\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\end{table}
The situations for classical quantities and the computational power
above them are summarized in Table~\ref{tab:Comparison}, and similar
situations could potentially be applied to quantum mechanics and computation.
The physical quantity manipulated by the quantum circuit model, the
most widely used model of quantum computing, is the probability amplitudes
of states. \todo{Introduction Cleaned up until here. }Because we
are agnostic about whether probability amplitude is quantized or can
really be valued in the continuum of complex numbers, we tried to
develop two different types of models: the quantum theories and computing
over finite fields are discrete models, and the quantum interval-valued
probability measures incorporate the idea of measurement precision
into the quantum theory. To distinguish the modified quantum theories
from the usual ones, the usual ones will be called the ``conventional
quantum theory'' where necessary and it will be reviewed in Chapter~\ref{chap:Conventional-Quantum-Theory}~\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.}.

@@

\centering{}%
\begin{tabular}{ccccccc}
\toprule 
\addlinespace
 & Conventional & Modal & Discrete (I) & Discrete (II) & QPMFF & QIVPM\tabularnewline\addlinespace
\midrule
\addlinespace
\midrule 
\addlinespace
 & Chapter~\ref{chap:Conventional-Quantum-Theory} & Sec.~\ref{modalquantum} & Sec.~\ref{discretequantumtheoryI} to \ref{sec:Geometry-of-Entangled-States} & Sec.~\ref{sec:Discrete-Quantum-Theory-(II)} & Sec.~\ref{sec:Toward-IVPM} & Chapter~\ref{chap:QIVPM}\tabularnewline\addlinespace
\midrule
\addlinespace
\addlinespace
States space & $\mathbb{C}^{D}$ & $\Fq^{D}$ & $\Fpp^{D}$ & %
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Local region\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
in $\Fpp^{D}$\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & $\Fpp^{D}$ & $\mathbb{C}^{D}$\tabularnewline\addlinespace
\addlinespace
\addlinespace
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Likelihood\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
of events is\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
predicted by\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & %
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Real-valued\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
probability\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & %
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Possible or\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
impossible\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & %
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Possible or\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
impossible\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & %
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Cardinal\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
probability\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & %
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Real-valued\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
probability, but no\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
sensible Born rule\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & %
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Interval-valued\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
probability\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular}\tabularnewline\addlinespace
\addlinespace
\addlinespace
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Expectation\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
value\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & Defined & Undefined & Undefined & Undefined &  & Defined\tabularnewline\addlinespace
\addlinespace
\midrule 
\addlinespace
 &  & Sec.~\ref{modalquantumcomputing} & Sec.~\ref{discretequantumcomputingI} & Sec.~\ref{discretequantumcomputingII} &  & \tabularnewline\addlinespace
\midrule
\addlinespace
\addlinespace
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Hadamard\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
transformation\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & Included & %
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
May not be\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
included\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & Included & Included &  & \tabularnewline\addlinespace
\addlinespace
\addlinespace
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Solve\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
$\usat$\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
efficiently\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & Unlikely & Yes & Partially & Unlikely &  & \tabularnewline\addlinespace
\addlinespace
\addlinespace
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Grover search\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
algorithm\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & Yes &  &  & Yes &  & \tabularnewline\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\end{sidewaystable}
\todo{Beginning of Merge Left}Since conventional probability amplitude
is in the field of complex numbers, and the field structure is essential
to compute the probability from some probability amplitudes, we in
Chapter~\ref{chap:DQT=000026DQC} tried to build discrete quantum
theories and computing by considering them over finite fields.

>>>>

\usepackage{tabularx}
\newcolumntype{Y}{>{\centering\doublespacing\arraybackslash}X}

@@

Since we as human cannot distinguish quantities differing only by
an arbitrarily small extent \cite{Turing_1937,Gisin2017}, we want
to incorporate this limitation into our quantum mechanical model to
better capture our ability of predicting quantum phenomenon and the
power of realistic quantum computers. Since we are agnostic about
whether the reality is ultimately discrete or continuous, we considered
two types of quantum models. By assuming the quantum states are ultimately
discrete and distinguishable, our first type of quantum theories replaces
the complex numbers by finite fields. We then focus on the limitation
of distinguishability itself, incorporate the finite precision measurement
into our model, and consider quantum interval-valued probability measures.

After first suggested by Richard Feynman 40 years ago~\cite{Feynman1982Simulating},
IBM, \textitalian{Rigetti}, Google, Microsoft, and many other companies
and countries start to build realistic quantum computers recently
\cite{Shieber2018,Kahn2019}, and seek more efficient ways to simulate
chemical molecules and to break down RSA \cite{boneh1999twenty,wiki:RSAFactoring}.
No matter how, classical or quantum, to attack these two tasks, they
require two different types of quantities. On one hand, the question
and answer of RSA are both discrete integers. On the other hand, even
if the quantities describing molecules are considered as continuous,
we want to argue that it is beyond our ability to decide these quantities,
including the coordinate of the wave functions, are ultimately discrete
or continuous. To better understand this issue without struggling
with their meaning in the quantum theory, it is instructive to review
how classical quantities are characterized and computed.

@@

In contrast, the non-apparently discrete quantities confused many
cultures since very beginning. For example, the ancient Chinese philosopher
Chuang-tzu (庄子) didn't believe these quantities have a minimum unit,
and expressed ``If a one-foot-long stick is cut into halves every
day, the cutting will never come to an end''\footnote{``一尺之棰，日取其半，萬世不竭。'' in Chinese \cite{Zhuangzi1999,Liu2018}.}.
As another example, Greek mathematicians in the school of Pythagoras
believed every mathematical model must be ultimately discrete, and
they in legend murdered Hippasus because he proved there is no minimum
unit between a side of a square and its diagonal. Comparing the idea
that the length of a square's diagonal is not a physical quantity
or there is no square physically, working with a mathematical model
without minimum unit seems more appealing. Since the mathematical
model is continuous, their corresponding physical quantities are naturally
assumed to be continuous. Later, these continuous quantities become
the basic component of analog computers. After logarithm was discovered,
a slide rule, also known as a slipstick, is used to compute multiplication,
division, and more complex operations as a mechanical analog computer~\cite{wiki:SlideRule}.
Charge amplifiers are used to build electrical analog computers and
compute the integration in calculus \cite{wiki:Integrator,wiki:ChargeAmplifier}.

Although it is convenient to assume physical quantities is as continuous
as their mathematical models, the physical quantities discussed in
previous paragraph may not simply represented by real numbers. Chuang-tzu's
stick are made by molecules and cannot be cut in halves endlessly.
The input charge of an integrator must actually be an integer multiple
of the elementary charge. A prefect square cannot be draw physically.
Even we don't draw it, but merely asking whether it exists in the
space is still beyond our ability to answer because whether the space
itself is ultimately discrete or continuous is still an unsolved question.
Even if it exists, it cannot be distinguished by human from an almost
prefect square differing only by an arbitrarily small extent. Similar
situation applying to the length of a slide rule, it is beyond our
ability to decide whether its length is ultimately discrete or continuous.
Even if it is continuous, identifying which real number represents
its length requires infinite precision, but the precision in reality
is limited generally three or four significant figures~\cite{wiki:AnalogComputer}.
As the results, even if analog computers really store real numbers,
they cannot be read, write, and use to its exact value to branch the
computation. The opposite of the last condition is sometimes assumed
by real computational models. For example, Blum, Cucker, Shub, and
Smale modeled their BCSS machine to allow the operations of deciding
whether a number is greater or equal to zero over real numbers and
whether a number is exactly zero over complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}.
Therefore, the computational power of analog computers is far weaker
than their theoretical prediction.

\begin{sidewaystable}
\caption{\label{tab:Comparison}Comparison between the classical quantities
and their mathematical models. \todo{Add the quantum case, and center
entries vertically.}}

\centering{}\centering{}%
\begin{tabularx}{\linewidth}{YcYY}
\toprule 
\addlinespace
Physical quantities & Discrete & Discrete with extreme small unit & Continuous?\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational devices & Digital computer & Analog computer & Analog computer\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Mathematical representation & Integers & Real numbers & Real numbers\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational model  & Turing machine & BCSS machine & BCSS machine\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
How the models realize? & Reliably & 
Not reliably, but could be represented by a faithful discrete model when the minimum unit could be manipulated with a better technology. & 
Not reliably, because precision can never be high enough. \tabularnewline\addlinespace
\bottomrule
\addlinespace
\end{tabularx}
\end{sidewaystable}
The situations for classical quantities and the computational power
above them are summarized in Table~\ref{tab:Comparison}, and similar
situations could potentially be applied to quantum mechanics and computation.
The physical quantity manipulated by the quantum circuit model, the
most widely used model of quantum computing, is the probability amplitudes
of states. \todo{Introduction Cleaned up until here. }Because we
are agnostic about whether probability amplitude is quantized or can
really be valued in the continuum of complex numbers, we tried to
develop two different types of models: the quantum theories and computing
over finite fields are discrete models, and the quantum interval-valued
probability measures incorporate the idea of measurement precision
into the quantum theory. To distinguish the modified quantum theories
from the usual ones, the usual ones will be called the ``conventional
quantum theory'' where necessary and it will be reviewed in Chapter~\ref{chap:Conventional-Quantum-Theory}~\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.}.

@@

\begin{tabularx}{\linewidth}{YYYYYYY}
\toprule 
\addlinespace
 & Conventional & Modal & Discrete (I) & Discrete (II) & QPMFF & QIVPM\tabularnewline\addlinespace
\addlinespace
\midrule
\midrule 
\addlinespace
 & Chapter~\ref{chap:Conventional-Quantum-Theory} & Sec.~\ref{modalquantum} & Sec.~\ref{discretequantumtheoryI} to \ref{sec:Geometry-of-Entangled-States} & Sec.~\ref{sec:Discrete-Quantum-Theory-(II)} & Sec.~\ref{sec:Toward-IVPM} & Chapter~\ref{chap:QIVPM}\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
States space & $\mathbb{C}^{D}$ & $\Fq^{D}$ & $\Fpp^{D}$ & Local region in $\Fpp^{D}$ & $\Fpp^{D}$ & $\mathbb{C}^{D}$\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Likelihood of events is predicted by & Real-valued probability & Possible or impossible & Possible or impossible & Cardinal probability & Real-valued probability, but no sensible Born rule & Interval-valued probability\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Expectation value & Defined & Undefined & Undefined & Undefined &  & Defined\tabularnewline\addlinespace
\addlinespace
\midrule
\midrule 
\addlinespace
 &  & Sec.~\ref{modalquantumcomputing} & Sec.~\ref{discretequantumcomputingI} & Sec.~\ref{discretequantumcomputingII} &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Hadamard transformation & Included & May not be included & Included & Included &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Solve $\usat$ efficiently & Unlikely & Yes & Partially & Unlikely &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Grover search algorithm & Yes &  &  & Yes &  & \tabularnewline\addlinespace
\bottomrule
\addlinespace
\end{tabularx}
\end{sidewaystable}
\todo{Beginning of Merge Left}Since conventional probability amplitude
is in the field of complex numbers, and the field structure is essential
to compute the probability from some probability amplitudes, we in
Chapter~\ref{chap:DQT=000026DQC} tried to build discrete quantum
theories and computing by considering them over finite fields.
